Design and Implementation of Probabilistic Programming Language Anglican

### 2016-07-04 paper/paper.tex
in progress

\begin{abstract}
    In these notes, I explain the rationale behind the
    macro-compiled implementation of Anglican, outline
    design choices, and describe important implementation
    aspects.
\end{abstract}

\section{Introduction}

@@

>>>>

\begin{abstract}
    Anglican is a probabilistic programming system designed to interoperate
    with Clojure and other JVM languages. We describe the implementation of
    Anglican and illustrate how its design facilitates both explorative and
    industrial use of probabilistic programming.

    In these notes, I explain the rationale behind the
    macro-compiled implementation of Anglican, outline
    design choices, and describe important implementation
    aspects.
\end{abstract}

\section{Introduction}
For data science practitioners, statistical inference is typically
but one step in a more elaborate analysis workflow. The first stage
of this work involves data acquisition, pre-processing and cleaning.
This is often followed by several iterations of exploratory model
design and testing of inference algorithms. Once a sufficiently
robust statistical model and corresponding inference algorithm have
been identified, analysis results must be post-processed, visualized,
and in some cases integrated into a wider production system.

Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
represent generative models as programs written in a specialized
language that provides syntax for the definition and conditioning of
random variables. The code for such models is generally concise,
modular, and easy to modify or extend. Typically inference can be
performed for any probabilistic program using one or more generic
inference techniques provided by the system backend, such as
Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
\cite{SDT04}, expectation propagation \cite{MWG+10}, and extensions
of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
Although these generic techniques are not always as statistically
efficient as techniques that take advantage of model-specific
optimizations, probabilistic programming makes it easier to optimize
models for a specific application in a manner that is efficient in
terms of the dimensionality of its latent variables.

While probabilistic programming systems shorten the iteration cycle in
exploratory model design, they typically lack basic functionality needed for
data I/O, pre-processing, and analysis and visualization of inference results.
In this demonstration, we describe the implementation of
Anglican~(\url{http://www.robots.ox.ac.uk/~fwood/anglican/}), a probabilistic
programming language that tightly integrates with
Clojure~(\url{http://clojure.org/}), a general-purpose programming language
that runs on the Java Virtual Machine (JVM). Both languages share a common
syntax, and can be invoked from each other. This allows Anglican programs to
make use of a rich set of libraries written in both Clojure and Java.
Conversely Anglican allows intuitive and compact specification of models for
which inference may be performed as part of a larger Clojure project.

@@

An Anglican program, or \textit{query}, is compiled into a Clojure
function. When inference is performed with a provided algorithm, this
produces a sequence of return values, or \textit{predicts}. Anglican
shares a common syntax with Clojure; Clojure functions can be called
from Anglican code and vice versa. A simple program in Anglican can
look like the following code:

%\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
%(defquery models
%  "chooses a distribution which describes the data"
%  (let [;; Model -- randomly choose a distribution and parameters
%        dist (sample (categorical [[uniform-discrete 1]
%                                   [uniform-continuous 1]
%                                   [normal 1] 
%                                   [gamma 1]]))
%        a (sample (gamma 1 1)) b (sample (gamma 1 1))
%        d (dist a b)]
%    ;; Data  --- samples from the unknown distribution
%    (observe d 1) (observe d 2) (observe d 4) (observe d 7)
%    ;; Output --- predicted distribution type and parameters
%    (predict :d (type d))
%    (predict :a a) (predict :b b)))
%\end{lstlisting}

Internally, an Anglican query is represented by a computation in
\textit{continuation passing style} (CPS), and inference algorithms
exploit the CPS structure of the code to intercept probabilistic
operations in an algorithm-specific way. Among the available
inference algorithms there are Particle Cascade~\cite{PWD+14},
Lightweight Metropolis-Hastings~\cite{WSG11}, Iterative Conditional
Sequential Monte-Carlo (Particle Gibbs)~\cite{WVM14}, and others.
Inference on Anglican queries generates a lazy sequence of
samples, which can be processed asynchronously in Clojure
code for analysis, integration, and decision making.

Clojure (and Anglican) run on the JVM and get access to a wide choice of Java
libraries for data processing, networking, presentation, and imaging.
Conversely, Anglican queries can be called from Java and other JVM languages.
Programs involving Anglican queries can be deployed as JVM
\textit{jars}, and run without modification on any platform for 
which JVM is available.
